---
title: "Episodes of Democratic Backsliding"
format: html
embed-resources: true
editor: source
execute:
  echo: true
  warning: false
  message: false
---

## Data

I use the [Episodes of Regime Transformation (ERT) dataset](https://www.v-dem.net/data/ert-dataset/) from the V-Dem Institute. The dataset identifies episodes of democratization and autocratization across countries over time.

```{r}
#| label: prepare-episodes
library(tidyverse)
ert <- read_csv("data/ert.csv")
# glimpse(ert)
# names(ert)

# Keep one row per autocratization episode: its first year
aut_start <- ert |>
  filter(
    aut_ep == 1,
    year == aut_ep_start_year
  ) |>
 mutate(
    regression_type = case_when(
      aut_ep_prch == 1 ~ "Democratic regression",
      aut_ep_prch == 0 ~ "Autocratic regression",
      TRUE ~ NA_character_
    ),
    outcome = case_when(
      aut_ep_outcome == 1 ~ "Democratic breakdown",
      aut_ep_outcome == 2 ~ "Preempted breakdown",
      aut_ep_outcome == 3 ~ "Diminished democracy",
      aut_ep_outcome == 4 ~ "Averted regression",
      aut_ep_outcome == 5 ~ "Regressed autocracy",
      aut_ep_outcome == 6 ~ "Outcome censored",
      TRUE ~ "Ongoing or unavailable"
    ),
    duration = aut_ep_end_year - aut_ep_start_year + 1
  ) |>
  filter(!is.na(regression_type)) |>
  distinct(aut_ep_id, .keep_all = TRUE)

```

## Visualization 1: The Changing Composition of Autocratization

```{r}
#| label: fig-autocratization-over-time
#| fig-width: 10
#| fig-height: 6

episodes_by_year <- aut_start |>
  count(aut_ep_start_year, regression_type, name = "n_episodes")

ggplot(
  episodes_by_year,
  aes(
    x = aut_ep_start_year,
    y = n_episodes,
    fill = regression_type
  )
) +
  geom_area(
    position = "stack",
    alpha = 0.9,
    colour = "white",
    linewidth = 0.2
  ) +
  scale_fill_manual(
    values = c(
      "Democratic regression" = "#C94C4C",
      "Autocratic regression" = "#4C78A8"
    ),
    name = "Episode type"
  ) +
  scale_x_continuous(breaks = seq(1900, 2025, by = 20)) +
  labs(
    title = "Autocratization Episodes Have Increased in Recent Decades",
    subtitle = "New episodes, separated by whether they began in a democracy or autocracy",
    x = "Episode start year",
    y = "Number of new autocratization episodes",
    caption = "Source: V-Dem Institute, Episodes of Regime Transformation (ERT) Dataset, v16. Author's calculations."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )

```

## Visualization 2: How Long Does Autocratization Last?

```{r}
#| label: fig-autocratization-duration
#| fig-width: 9
#| fig-height: 6

ggplot(
  aut_start,
  aes(
    x = regression_type,
    y = duration,
    fill = regression_type,
    colour = regression_type
  )
) +
  geom_boxplot(
    width = 0.55,
    outlier.shape = NA,
    alpha = 0.65
  ) +
  geom_jitter(
    width = 0.12,
    height = 0,
    size = 2,
    alpha = 0.45
  ) +
  scale_fill_manual(
    values = c(
      "Democratic regression" = "#F4A6A6",
      "Autocratic regression" = "#A9C7E8"
    ),
    name = "Episode type"
  ) +
  scale_colour_manual(
    values = c(
      "Democratic regression" = "#B22222",
      "Autocratic regression" = "#1F5A91"
    ),
    guide = "none"
  ) +
  coord_flip() +
  labs(
    title = "Autocratization Episodes Vary Widely in Duration",
    subtitle = "Each point represents one autocratization episode",
    x = NULL,
    y = "Episode duration (years)",
    caption = "Source: V-Dem Institute, Episodes of Regime Transformation (ERT) Dataset, v16. Author's calculations."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )
```

## Visualization 3: Democratic Trajectories after Backsliding Begins

#### It compares democracy scores in the five years before and ten years after an autocratization episode begins.

```{r}
#| label: fig-democracy-trajectories
#| fig-width: 10
#| fig-height: 6

trajectory_episodes <- aut_start |>
  filter(
    regression_type == "Democratic regression",
    outcome %in% c("Democratic breakdown", "Averted regression")
  ) |>
  select(
    country_text_id,
    episode_start = aut_ep_start_year,
    outcome
  )

trajectory_data <- trajectory_episodes |>
  inner_join(
    ert |>
      select(country_text_id, year, v2x_polyarchy),
    by = "country_text_id"
  ) |>
  mutate(relative_year = year - episode_start) |>
  filter(between(relative_year, -5, 10))

trajectory_summary <- trajectory_data |>
  group_by(relative_year, outcome) |>
  summarise(
    mean_polyarchy = mean(v2x_polyarchy, na.rm = TRUE),
    se = sd(v2x_polyarchy, na.rm = TRUE) / sqrt(sum(!is.na(v2x_polyarchy))),
    lower = pmax(0, mean_polyarchy - 1.96 * se),
    upper = pmin(1, mean_polyarchy + 1.96 * se),
    .groups = "drop"
  )

ggplot(
  trajectory_summary,
  aes(
    x = relative_year,
    y = mean_polyarchy,
    colour = outcome,
    fill = outcome,
    linetype = outcome,
    group = outcome
  )
) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.15,
    colour = NA
  ) +
  geom_line(linewidth = 1.1) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    colour = "grey35",
    linewidth = 0.7
  ) +
  scale_colour_manual(
    values = c(
      "Democratic breakdown" = "#B22222",
      "Averted regression" = "#238B45"
    ),
    name = "Episode outcome"
  ) +
  scale_fill_manual(
    values = c(
      "Democratic breakdown" = "#B22222",
      "Averted regression" = "#238B45"
    ),
    name = "Episode outcome"
  ) +
  scale_linetype_manual(
    values = c(
      "Democratic breakdown" = "solid",
      "Averted regression" = "dashed"
    ),
    name = "Episode outcome"
  ) +
  labs(
    title = "Backsliding Has Different Democratic Trajectories",
    subtitle = "Average electoral-democracy score before and after autocratization begins; shaded bands are 95% confidence intervals",
    x = "Years relative to autocratization onset",
    y = "Average electoral-democracy score",
    caption = "Source: V-Dem Institute, Episodes of Regime Transformation (ERT) Dataset, v16. Author's calculations."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "top"
  )
```

## Visualization 4: Recent Autocratization Episodes by Country

```{r}
#| label: fig-recent-episode-heatmap
#| fig-width: 10
#| fig-height: 8

recent_episodes <- aut_start |>
  arrange(desc(aut_ep_start_year)) |>
  slice_head(n = 20) |>
  select(aut_ep_id, outcome)

recent_heatmap <- ert |>
  filter(aut_ep == 1) |>
  inner_join(recent_episodes, by = "aut_ep_id") |>
  mutate(
    episode_label = paste0(country_name, " (", aut_ep_start_year, ")"),
    episode_label = fct_reorder(
      episode_label,
      aut_ep_start_year,
      .desc = TRUE
    )
  )

ggplot(
  recent_heatmap,
  aes(
    x = year,
    y = episode_label,
    fill = v2x_polyarchy
  )
) +
  geom_tile(
    colour = "white",
    linewidth = 0.15
  ) +
  scale_fill_viridis_c(
    option = "C",
    name = "Electoral-democracy\nscore"
  ) +
  labs(
    title = "Twenty Most Recent Autocratization Episodes",
    subtitle = "Tile color shows each country's electoral-democracy score during the episode",
    x = "Year",
    y = "Country and episode start year",
    caption = "Source: V-Dem Institute, Episodes of Regime Transformation (ERT) Dataset, v16. Author's calculations."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 8)
  )
```
